The Categorical Structure of Profunctor Optics

### 2017-07-21 paper.tex
Thinking about semioptics

\todo{Introduction should include:
\begin{itemize}
\item Previous attempts to generalise optics
\item ...
\end{itemize}
}

@@

There is a well known formula for the coend as the coequaliser in the diagram
\[
\begin{tikzcd}
\coprod_{M \to N} S(N, M) \ar[r,shift left=.75ex]  \ar[r,shift right=.75ex] & \coprod_{M \in \M} S(M, M) \ar[r] & \int^{M \in \M} S(M, M)
\end{tikzcd}
\]
where $S$ is some functor $\M^\op \times \M \to \E$.

@@

\begin{proposition}
If $\langle l, r \rangle : S \hto A$ is a lens then the associated $\fget$ and $\fput$ functions satisfy the $\fget\fput$, $\fput\fget$ and $\fput\fput$ laws.
\end{proposition}
\begin{proof}
\todo{Straightforward, written out at \cite{IsomorphismLensesPost}, definitely not the origin.}
\end{proof}

@@

\begin{proposition}
Suppose $\C$ has finite limits and that there is a morphism $x : 1 \to A$. Then if $\fget : S \to A$ and $\fput : S \times A \to S$ satisfy the three lens laws then the corresponding $p : S \hto A$ is a lens.
\end{proposition}
\begin{proof}
Set $C$ to be the pullback of $\fget$ along $x$, so there is a map $i : C \to S$ with $\fget \, i = x !_C$. There is also a map $j : S \to C$ induced by the following diagram:
\[
\begin{tikzcd}
S \ar[ddr, bend right = 20] \ar[dr, "j", dashed] \ar[r, "{[\id_S, x !_S]}"] & S \times A \ar[dr, "\fput"] & \\
& C \ar[r, "i"] \ar[d] \arrow[dr, phantom, "\lrcorner", very near start] & S \ar[d, "\fget"] \\
& 1\ar[r, "x", swap] & A
\end{tikzcd}
\]
which commutes by the $\fput\fget$ law. Note that $ji = \id_C$ by the universal property of pullbacks. Now the commutative diagram
\[
\begin{tikzcd}
C \times A \ar[r, "i \times A"] \ar[d, "\fput (i \times A)", swap] & S \times A \ar[d, "\fput"] \\
S \ar[r, equal] \ar[d, "{[j,\fget]}", swap] & S \ar[d, "{[\id_S, \fget]}"]  \\
C \times A \ar[r, "i \times A", swap] & S \times A
\end{tikzcd}
\]
is a witness that $\langle \fput (i \times A), [j,\fget] \rangle = \langle \fput, [\id_S, \fget] \rangle$ as elements of $\PreOptic_\M((S, S), (A, A))$. It remains to show that the maps on the left are mutual inverses:
\begin{align*}
\fput (i \times A)[j,\fget] &= \fput [ij,\fget] \\
&= \fput [\fput [\id_S, x!_S],\fget] && \text{(by definition of $j$)} \\
&= \fput [\id_S,\fget] && \text{(by $\fput\fput$)} \\
&= \id_S && \text{(by $\fget\fput$)}
\intertext{and}
[j,\fget]\fput (i \times A) &= [j\fput (i \times A),\fget\,\fput (i \times A)] && \text{(by universal property of product)} \\
&= [j\fput (i \times A), \pi_2 (i \times A)] && \text{(by $\fput\fget$)} \\
&= [j\fput (i \times A), \pi_2] && \\
&= [jij\fput (i \times A), \pi_2] && \\
&= [j\fput [\id_S,x !_S] \fput (i \times A), \pi_2] && \\
&= [j\fput [\id_S, x !_S] \pi_1 (i \times A), \pi_2] && \text{(by $\fput\fput$ \todo{todo: detail})}\\
&= [jij \pi_1 (i \times A), \pi_2] && \\
&= [jiji \pi_1, \pi_2] && \\
&= [\pi_1, \pi_2] && \\
&= \id_{C \times A}
\end{align*}
Therefore $\langle \fput (i \times A), [j,\fget] \rangle$ is an optic, so  $\langle \fput, [\id_S, \fget] \rangle$ is an optic, as we were hoping to show. \todo{maybe do the above in pictures}
\end{proof}

@@

\begin{proposition}
A pre-lens $p : S \hto A$ obeys the $\fput\fget$ and $\fget\fput$ laws iff $outside(p) = \id_S$ and $inside(p) = \langle t \times A \rangle$.
\end{proposition}
\begin{proof}
\todo{I haven't actually checked this yet}
\end{proof}

@@

Past attempts \cite{ReflectionsOnMonadicLenses} to define effectful lenses have wrapped the result of one or both of $\fget$ and $\fput$ with a monad. This seems to me a bit ad-hoc; it is not obvious what the laws ought to be, and there is no clear generalisation to other varieties of optic. The definition of optic given in this paper suggests we instead set $\C$ to be the Kleisli category for some monad.

>>>>

\todo{Introduction should include:
\begin{itemize}
\item Previous attempts to generalise optics
\item ...
\end{itemize}
}
A lens from $S$ to $A$ is a pair of maps $\fget : S \to A$ and $\fput : S \times A \to S$ that typically obeys three laws (whose names are in diagrammatic order):
\begin{itemize}
\item ``$\fput\fget$'': $\fget \; \fput = \pi_2$ as maps $S \times A \to A$, so that any update to $A$ should be \todo{represented faithfully in $S$},
\item ``$\fget\fput$'': $\fput [\id_S, \fget] = \id_S$ as maps $S \to S$, so that if $A$ is not changed then $S$ should not change; and,
\item ``$\fput\fput$'': $\fput (\fput \times \id_A) = \fput \, \pi_{1, 3}$ as maps $S \times A \times A \to S$, so any update to $A$ completely overwrites previous updates.
\end{itemize}

@@

There is a well known formula for the coend as the coequaliser in the diagram
\[
\begin{tikzcd}
\coprod_{M \to N} P(N, M) \ar[r,shift left=.75ex]  \ar[r,shift right=.75ex] & \coprod_{M \in \M} P(M, M) \ar[r] & \int^{M \in \M} P(M, M)
\end{tikzcd}
\]
where $S$ is some functor $\M^\op \times \M \to \E$.

@@

\begin{proposition}
\label{prop-OpticImpliesLensLaws}
If $\langle l, r \rangle : S \hto A$ is a lens then the associated $\fget$ and $\fput$ functions satisfy the $\fget\fput$, $\fput\fget$ and $\fput\fput$ laws.
\end{proposition}
\begin{proof}
Let $\fget = \pi_2 l$ and $\fput = r (\pi_1 l \times A)$. Verification of the laws is straightforward:
\begin{align*}
\fget \; \fput 
&= \pi_2 l r (\pi_1 l \times A) \\
&= \pi_2 (\pi_1 l \times A) \\
&= \pi_2 \\
\fput [\id_S, \fget] 
&= r (\pi_1 l \times A) [\id_S, \pi_2 l] \\
&= r [\pi_1 l, \pi_2 l] \\
&= rl \\
&= \id_S \\
\fput (\fput \times A) 
&= r (\pi_1 l \times A) (r (\pi_1 l \times A) \times A) \\
&= r (\pi_1 l  r (\pi_1 l \times A) \times A) \\
&= r (\pi_1 (\pi_1 l \times A) \times A) \\
&= r (\pi_1 l \times A) \pi_{1,3} \\
&= \fput \, \pi_{1, 3}
\end{align*}

\todo{Written out at \cite{IsomorphismLensesPost}, definitely not the first place though.}
\end{proof}

@@

\begin{proposition}
Suppose $\C$ has finite limits and that there is a morphism $x : 1 \to A$. Then if $\fget : S \to A$ and $\fput : S \times A \to S$ satisfy the three lens laws then the corresponding $p : S \hto A$ is a lens.
\end{proposition}
\begin{proof}
Set $C$ to be the pullback of $\fget$ along $x$, so there is a map $i : C \to S$ with $\fget \, i = x !_C$. There is also a map $j : S \to C$ induced by the following diagram:
\[
\begin{tikzcd}
S \ar[ddr, bend right = 20] \ar[dr, "j", dashed] \ar[r, "{[\id_S, x !_S]}"] & S \times A \ar[dr, "\fput"] & \\
& C \ar[r, "i"] \ar[d] \arrow[dr, phantom, "\lrcorner", very near start] & S \ar[d, "\fget"] \\
& 1\ar[r, "x", swap] & A
\end{tikzcd}
\]
which commutes by the $\fput\fget$ law. Note that $ji = \id_C$ by the universal property of pullbacks. Now the commutative diagram
\[
\begin{tikzcd}
C \times A \ar[r, "i \times A"] \ar[d, "\fput (i \times A)", swap] & S \times A \ar[d, "\fput"] \\
S \ar[r, equal] \ar[d, "{[j,\fget]}", swap] & S \ar[d, "{[\id_S, \fget]}"]  \\
C \times A \ar[r, "i \times A", swap] & S \times A
\end{tikzcd}
\]
is a witness that $\langle \fput (i \times A), [j,\fget] \rangle = \langle \fput, [\id_S, \fget] \rangle$ as elements of $\PreOptic_\M((S, S), (A, A))$. It remains to show that the maps on the left are mutual inverses:
\begin{align*}
\fput (i \times A)[j,\fget] &= \fput [ij,\fget] \\
&= \fput [\fput [\id_S, x!_S],\fget] && \text{(by definition of $j$)} \\
&= \fput [\id_S,\fget] && \text{(by $\fput\fput$)} \\
&= \id_S && \text{(by $\fget\fput$)}
\intertext{and}
[j,\fget]\fput (i \times A) &= [j\fput (i \times A),\fget\,\fput (i \times A)] && \text{(by universal property of product)} \\
&= [j\fput (i \times A), \pi_2 (i \times A)] && \text{(by $\fput\fget$)} \\
&= [j\fput (i \times A), \pi_2] && \\
&= [jij\fput (i \times A), \pi_2] && \\
&= [j\fput [\id_S,x !_S] \fput (i \times A), \pi_2] && \\
&= [j\fput [\id_S, x !_S] \pi_1 (i \times A), \pi_2] && \text{(by $\fput\fput$ \todo{todo: detail})}\\
&= [jij \pi_1 (i \times A), \pi_2] && \\
&= [jiji \pi_1, \pi_2] && \\
&= [\pi_1, \pi_2] && \\
&= \id_{C \times A}
\end{align*}
Therefore $\langle \fput (i \times A), [j,\fget] \rangle$ is a lens, so  $\langle \fput, [\id_S, \fget] \rangle$ is a lens as we were hoping to show. \todo{maybe do the above in pictures}
\end{proof}

@@

\begin{example}
A minimal example is the following:
\begin{align*}
\fget : \{A,B,C\} &\to \{X, Y\} \\
A &\mapsto X \\
B &\mapsto X \\
C &\mapsto Y 
\end{align*}
\begin{align*}
\fput : \{A,B,C\} \times \{X, Y\} &\to \{A,B,C\} \\
(A,X) &\mapsto A \\
(B,X) &\mapsto B \\
(C,X) &\mapsto B \\
(A,Y) &\mapsto C \\
(B,Y) &\mapsto C \\
(C,Y) &\mapsto C
\end{align*}
\end{example}

\todo{THE FOLLOWING IS TOTALLY WRONG, does not capture the right thing, fails for the above example}

\begin{proposition}
Suppose the pre-lens $\langle l, r \rangle : S \hto A$ satisfies $rl = \id_S$ and $\langle lr \rangle = \langle t \times A \rangle$ as an element of $\int^{M \in \M} \C(M \times A, M \times A)$. Then $p$ obeys the $\fput\fget$ and $\fget\fput$ laws.
\end{proposition}
\begin{proof}
This is nearly identical to Proposition \ref{prop-OpticImpliesLensLaws}. As was mentioned in the proof of Proposition \ref{prop-onthenose}, we may assume that $lr = t \times A$ on the nose.

Now, again letting $\fget = \pi_2 l$ and $\fput = r (\pi_1 l \times A)$,
\begin{align*}
\fget \; \fput 
&= \pi_2 l r (\pi_1 l \times A) \\
&= \pi_2 (t \times A) (\pi_1 l \times A) \\
&= \pi_2 (t \pi_1 l \times A) \\
&= \pi_2 \\
\fput [\id_S, \fget] 
&= r (\pi_1 l \times A) [\id_S, \pi_2 l] \\
&= r [\pi_1 l, \pi_2 l] \\
&= rl \\
&= \id_S
\end{align*}
\end{proof}

\begin{proposition}
The converse is true.
\end{proposition}
\begin{proof}

\end{proof}

@@

Past attempts \cite{ReflectionsOnMonadicLenses} to define effectful lenses have wrapped the result of one or both of $\fget$ and $\fput$ with a monad. This approach is a little ad-hoc; it is not obvious what the laws ought to be and there is no clear generalisation to other varieties of optic. The definition of optic given in this paper suggests we instead set $\C$ to be the Kleisli category for some monad.
